\documentclass[letterpaper,12pt,fullpage]{article}

\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
% \usepackage{cite}
\usepackage{graphicx}
% \usepackage[dvips]{graphicx}
% \usepackage{epsfig} % for postscript graphics files
  % \graphicspath{{../eps/}}
% \DeclareGraphicsExtensions{.eps}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[cmex10]{amsmath}
%\usepackage{array}
%\usepackage{mdwmath}
%\usepackage{mdwtab}
%\usepackage{eqparbox}
\usepackage[tight,footnotesize]{subfigure}
%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
%\usepackage{fixltx2e}
%\usepackage{stfloats}
\usepackage{hyperref}
\usepackage[backend=biber]{biblatex}
\addbibresource{exos/bleex.bib}
\addbibresource{exos/hal.bib}
\addbibresource{exos/xor.bib}
\addbibresource{exos/bodyExt.bib}
\addbibresource{exos/dualMod.bib}
\addbibresource{exos/ihmc.bib}
\addbibresource{exos/mit.bib}
\addbibresource{exos/roboKnee.bib}


% correct bad hyphenation here
%\hyphenation{op-tical net-works semi-conduc-tor}
\input{latex-commands}

\newcommand{\vF}{{\bf F}}
\newcommand{\vV}{{\bf V}}
\newcommand{\vG}{{\bf G}}
\newcommand{\vT}{{\bf T}}
\newcommand{\vR}{{\bf R}}


\begin{document}

\title{A Survey of Current Exoskeletons\\ and Their Control Architectures and
Algorithms\\
(Draft 3.2)}

\author{Alex Ansari, Christopher G. Atkeson, Howie Choset, and Matthew Travers\\
Carnegie Mellon University}

\maketitle

\begin{abstract}
In the simplest case, where the environment around a robotic system remains relatively fixed, controlling the dynamic interaction between the robot and the environment to achieve a specific physical objective is difficult.  In the case where an operator effectively wears the robot, i.e., a robotic exoskeleton, and would like to move through unknown and dynamically evolving environments, the situation becomes very complicated; the robotic system needs to account for the dynamics of multiple fundamentally linked systems that are, for the most part, difficult to model.  The goal of this document is to provide a survey of the literature on the design, modeling, and control of robotic exoskeletons and to highlight what we believe are several of the most promising currently existing approaches to the complicated goal of enabling dynamic yet safe interaction between operators, exoskeletons, and the environment.  Where information was available, we summarize specific contributions related to actuation, power storage and generation, sensor type and distribution, mechanical architecture, as well as control system design for a number of different exoskeleton hardware systems.  Included as part of this report are recommendations on what we see as helpful components as well as potentially problematic issues in current systems and control approaches. 
\end{abstract}

\section{Executive Summary}

This paper surveys the following exoskeletons in detail:
\begin{itemize}
\item
{\bf BLEEX:} a ``minimal-sensing'' and aggressive model-based control design that has influenced many subsequent exoskeletons.
\item
{\bf HAL:} an EMG controlled exoskeleton, emphasizing recognizing operator intent.
\item
{\bf XoR:} a hybrid Pneumatic/Electric-motor system, emphasizing the
value of using multiple actuator types.
\item
{\bf Body Extender:} a fully actuated full body exoskeleton. This is the
Iron Man option.
\item
{\bf HLEE:} a nice and well presented example of a torque control approach.
\item
{\bf IHMC MAE:} an SEA-based torque controlled exoskeleton, emphasizing
the value of mixing passive mechanisms (local energy storage) and active
actuation.
\item
{\bf MIT Exoskeleton:} an attempt to take advantage of the passive dynamics
of walking.
\item
{\bf RoboKnee:} an example of a single-joint exoskeleton. Powerful single-joint
or minimal exoskeletons are one way to enable operators to move fast
and naturally, and greatly increase their strength.
\end{itemize}

\noindent Exoskeletons in a variety of categories are listed in the Appendix.\\

\noindent Based on our review, we have made the following conclusions:
\begin{itemize}
\item Most exoskeletons estimate user intent indirectly, from measurements of suit variables.
\item Due to limitations in current sensory systems, it is difficult to sense user intent directly. 
\item As direct measurement of user intent is critical in dynamic, noisy environments, we recommend better fusion of low information density sensory data obtained from users with closed-loop control based on suit variables.
\item In the longer term, we recommend investigation and development of more advanced human-exoskeleton sensing technology to capture user intent.
\end{itemize}

\section{Introduction}

This document summarizes what we have found to be several of the most relevant exoskeleton hardware systems and the associated controllers used to enable the hardware to effectively and safely interact with human operators.  As we further elaborate, the different systems that we summarize range from the relatively simple and dramatically underactuated single joint mechanisms to extremely complex and fully actuated whole-body solutions.  The controllers associated with these systems also ranged from solutions that implemented conceptually simple designs to those which made use of novel and more complex design techniques and/or relatively untried sensing and actuation technologies. 

While this document does try to provide a relatively wide overview of existing exoskeleton technology, we did focus our efforts in this report on hardware systems meant to provide power augmentation and magnification as opposed the somewhat larger subset of exoskeletons designed primarily for rehabilitation.  This choice was made relative to both hardware as well as control system design considerations.  Exoskeletons designed for rehabilitation are not typically designed as stand-alone systems and are generally limited in the scope of operational modes in which they are capable of reliably providing assistance.  The hardware designed for athletic operators and extreme tasks will need to be high-powered with a power source that is capable of being carried such that the suit is a completely stand alone unit and will need to run a control system that is capable of providing transparent support to the operator in a large variety of operational scenarios.

To this end, the hardware systems that we chose to include summaries of in this document were the BLEEX, HAL, XoR, Body Extender, Hydraulic Lower Extremety, IHMC Mobility Assist, MIT, and RoboKnee exoskeletons.  The different systems cover many of the possible choices of actuation strategies, ranging from electric motors, to hydraulics, to pneumatics, and combinations thereof.  Several of the systems used novel transmission strategies as well, focusing on both efficient as well as high-fidelity force control implementations.  The transmission strategies included standard hydraulic systems, hydraulic systems with actively controlled bypass valves, ball-screw mechanisms that include series elastic components, as well as pneumatically actuated pulley-tendon and belt-drive mechanisms. 

In terms of sensing and control technologies, the different exoskeletons covered in this report employed a wide variety of the spectrum of available options.  Sensors were used to measure everything from standard information, like angular displacement of the hardware system's joints, to more novel information, like muscle activity of the user using EMG technology.  Sensor information was used as the basis for a variety of different closed and open-loop control system designs, ranging from straight-forward force control and trajectory following to novel approaches of fully actuated feedback linearization and sensitivity amplification strategies.  

Each of these component technologies are reviewed as they relate to the hardware systems discussed in this report.  Recommendations on what we feel are the best approaches in these reviews are also provided.  We provide a summary of these recommendations at the end of this document.

%\section{Scope: What is this paper about?}
%
%This paper surveys exoskeleton control architectures and
%algorithms.
%The focus is on exoskeleton control that allows a
%highly trained and top percentile athletic 
%operator to carry a payload that weighs approximately the same amount
%as the operator. We envisage these types of exoskeletons to be useful
%in carrying protective and safety equipment for SWAT teams, police,
%firefighters, and soldiers. 
%We also 
%focus this survey on exoskeleton control for lower body tasks (standing, walking,
%running, jumping, kicking, dodging, ...).

\input{exos/bleex}

\input{exos/hal}

\input{exos/xor}

\input{exos/exosText_bodyExt}

\input{exos/exosText_dualMode}

\input{exos/exosText_ihmc}

\input{exos/exosText_MIT}

\input{exos/exosText_roboKnee}

\section{Conclusions and Recommendations}
\label{survey:recommend}

The (strength augmentation) exoskeletons we reviewed can be categorized into two general control categories.  The exoskeleton either directly measures user intent (e.g., from EMG data or direct human-exoskeleton interaction forces), or it estimates intent from measurements at the exoskeleton joints and applies a sensitivity amplification strategy to shadow the wearer.  

Directly measuring user intent is difficult to implement using standard sensing modalities.  For instance, EMG data is notoriously noisy and presents both modeling and calibration challenges.  It would also be extremely difficult to keep these sensors in place to obtain accurate readings in dynamic environments or combat scenarios.  For similar reasons, many systems avoid the use of load cells or complicated sensor networks to accurately obtain operator-robot interaction forces / torques.  

Sensitivity amplification strategies circumvent many of the direct operator-robot sensing issues by estimating human intent from measurements of the exoskeleton hardware. The issue with this policy is that it cannot generally distinguish between user generated torques and those produced by external sources.  As such, they end up amplifying not only the user's intent, but also external disturbances, requiring the operator to expend valuable energy to stabilize the system.  
In dynamic situations the operator may not be capable of actively stabilizing the exoskeleton.
% Additionally, even under the assumption that external forces can be minimized or mitigated by the user, enabling the suit to be highly responsive to perceived user intent implies that the hardware needs to be operating near the margin of stability.  This is not an optimal strategy for highly dynamic behaviors, as the suit may be pushed beyond the point to which the operator can reasonably stabilize it.

% Based on these observations, our main recommendation is that a \emph{modified get out of the way} control strategy provides a reasonable policy for dynamic exoskeletons.
Based on these observations, we recommend (at least in the short term) fusing the two predominant approaches to exoskeleton control. That is, combining sensitivity amplification strategies with direct user measurement to robustly distinguish user intent from external disturbances.
% For instance, the get out of the way controller used in the Hydraulic Lower Extremity Exoskeleton would provide a starting point for this strategy.  This controller utilizes IMUs, foot contact sensors, and joint torque sensors to implement a user-reactive gravity compensating strategy that does not directly measure the user.  Augmenting the sensor suite with sensors that that measure user generated forces could allow this system to identify and reject external disturbances while simultaneously shadowing user motion.
%
% We emphasize that user intent is critical, and current sensing challenges must be addressed in order to develop dynamic exoskeletons capable of serving in combat scenarios.  As a stop-gap, it may be possible to combine low information density EMG (thresholded on/off muscle activation signals) and human force/torque or angular measurements to filter user intent.  Such an approach may prove useful in a hybrid strategy, filtering external disturbance forces during sensitivity amplification control.  
% As a longer term solution, we highlight research in the area of MEMs and nano-fabrication, which has lead to the development of new, flexible ``robotic skin'' sensing arrays.  Integrating these sensing arrays into operator clothing (or attaching directly to the exoskeleton), it may be possible to directly measure human interaction forces (i.e., estimating pressure / force and exoskeleton contact locations), which was not feasible in the recent past.  Dense sensing arrays add redundancy and can improve measurement reliability. We recommend further study and the development of more advanced sensory systems to address the challenges inherent in dynamic exoskeleton control.
%
For instance, it may be possible to use low information density EMG (thresholded on/off muscle activation signals) with robot force/torque and angular measurements to filter disturbances from operator-robot interaction forces.  As a longer term solution, we highlight research in the area of MEMs, nano-fabrication, soft sensors, etc., which has lead to the development of new, flexible ``robotic skin'' sensing arrays.  Integrating these sensing arrays into operator clothing (or attaching directly to the exoskeleton), it may be possible to directly measure human interaction forces (i.e., estimating pressure / force and exoskeleton contact locations), which was not feasible in the recent past.  Dense sensing arrays add redundancy and can improve measurement reliability and, thereby, feedback control performance.

We also highly stress the importance of high-bandwidth control loops.  It is important that an exoskeleton's control bandwidth stays well above (typically $2-3\times$) the human control bandwidth ($2-5\,$Hz), to track dynamic motion.  Several of the reviewed exoskeletons controllers are close to lower limit of this human bandwidth threshold under no-load conditions, and so can interfere in dynamic tasks.  Note that these bandwidth issues are primarily a function of the exoskeleton hardware, and thus special care/analysis needs to be given/conducted to make sure that the hardware is capable of physically producing required behaviors. 

Lastly, the importance of producing hardware that is physically capable of achieving dynamic performance is fundamentally linked to how mass is distributed throughout the system.  In general, it is desirable to minimize mass, particularly in distal regions. The design objective is often achieved by focusing actuation and leaving certain degrees of freedom passive (leading to an underactued system).  Underactuted control is a complicated problem that is often ignored in the sense that controllers assume the operator will provide necessary forces and torques for the un-actuated degrees of freedom.  Though it is possible to deal with the underactuated control problem directly (e.g., optimization-based controllers), there is a balance between feedback and feedforward control.  
% The better one is, the less reliant the system is on the other.  
By developing better sensors and feedback controllers that directly measure and shadow operator intent, exoskeleton controllers will become more robust and less sensitive to errors due to underactuated system assumptions and dynamic model inaccuracies.

%
%, or can be partially compensated using passive energy storage and dissipation elements.  
% we recommend a combination of biomechanically-inspired passive elements and optimization-based control methods, as these should ultimately produce the best strategy for dealing with underactuation in highly-dynamic exoskeleton systems.


%Such feedforward, model-based controllers depend heavily on accurate exoskeleton / human dynamics and so modeling efforts must be emphasized.

% Finally, we note that 
% Many of the common model-based control methods in this survey employ feedback linearization and inverse dynamics strategies for dynamic exoskeleton control.  
% These methods (incorrectly) assume the exoskeleton is fully actuated and require users to provide certain torques at uncontrolled joints.  There are more advanced control schemes that extend to underactuated systems and may apply here (e.g., optimization-based controllers or virtual model control).  However, we note that there is a balance between feedback and feedforward control.  The better one is, the less reliant the system is on the other.
%
% Improved feedback and direct sensing of user intent / human interaction forces therefore also offers to reduce sensitivity to high accuracy models and the fully actuated assumptions in inverse dynamics control. 

%% OTHER POINTS:
% Model-based approaches are heavily reliant on accurate models 
% Underactuated -- feedback linearization typically assumes fully actuated systems.
% human bandwidth vs control bandwidth

\newpage

\input{survey-atkeson}

\end{document}


Notes:

See if the Hydraulic Lower Extremity Exo is sensitivity amp and if so change opinion so we don't favor the approach.

Talk a lot about drawbacks of SEAs in IHMC.  Are these included in MIT / RoboKnee?  (low control bandwidth is mentioned in RoboKnee).  
We should have a blurb about the strengths/weakness of SEAs and possible options -- stiffer springs for force measurement and spring emulation for compliant control? 

Add references for new sensing tech
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
